Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-14 text/manuscript-full-KS.qmd
update appendix

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```
The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family income greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

[^5]: For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.

[^6]: Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer colleges, information that can be discerned from the Institutional Profiles (Appendix @tbl-sample-institutional-profile). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending behavior. Drawing from @RN4975, commodification of the Market Segment Model and Geomarkets engineers homophily.

@@

[^10]: Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket." In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago \[ADD TO APPENDIX\], of Lake Forest College, of Saint Mary's College \[ADD TO APPENDIX CRYSTAL\]

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings [@RN4011]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to make generalizations about Geomarket variation nationally, but to explore how different geographical models of metropolitan areas, shaped by their unique demographic characteristics, influence the composition of prospect pools. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomena in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) presents results for a larger number of metropolitan areas.

@@

Results are presented below for Chicago, Dallas, and Los Angeles. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) provides a more convenient means of viewing results. The online appendix also contains interactive maps, additional figures and tables not contained in the main text, and results for additional metropolitan areas (e.g., Bay Area, Philadelphia). The table of contents for navigating between metro areas can be accessed by clicking on the three horizontal lines on the bottom left of the screen.

@@

The interactive maps in the [Online Appendix](https://ozanj.github.io/cb_geo/#/research-question-1-15) can show these shifts at the census tract-level. For example, in 1980 Black people were concentrated in southern tracts of the City of Chicago Geomarket, as described by scholarship on gentrification in Chicago [@ewing2024_nofuture; @snidal2022_exodus]. However, in 2000 and 2020 Black populations shift towards northeast tracts of the South and Southwest Suburbs Geomarket

@@

These contribution patterns by race/ethnicity demonstrate that student list orders targeting prospects in the Chicago metropolitan area using Geomarket filters that exclude the City of Chicago -- as observed in some orders collected via our public requests -- result in the loss of substantially disproportionate shares of Black and Hispanic students from the resulting prospect pools. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) shows that this pattern holds across both lower-range (1020-1150) and higher-range (1310-1600) SAT score filters. In both cases, the City of Chicago contributes disproportionately larger shares of Black and Hispanic prospects compared to all other Geomarkets. <!--# [REVISE THESE SENTENCES TO HAMMER THE LEDE -- IF WE EXCLUDE CITY OF CHICAGO GEOMAREKT, THE WAY THE ASU ORDER DOES, WE LOSE A DISPROPORTIONATE SHARE OF BLACK AND HISPANIC STUDENTS] -->

@@

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles (11 Geomarkets) -- results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), results for which are shown in the [Online Appendix](https://ozanj.github.io/cb_geo/#/results). The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core [@Harris1945]. As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up of the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernardino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

@@

The Market Segment Model begins with the status attainment model idea that student demand for college is a function of parental education and parental income. @RN4982 illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that households of specific class characteristics and college aspirations are likely to live in particular communities and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard output in EPS software. When considering new Geomarkets to target, EPS recommends colleges utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A).

@@

\clearpage

\newpage

# Appendix A

```{r market-segment-characteristics, echo = F}
#| label: tbl-market-segment-characteristics
#| tbl-cap: 'Simplified market segment profile, Connecticut Market 3: Fairfield County'

market_segment <- data.frame(
  activity = c('Total test takers','Avg SAT (verbal + math)','Avg # scores sent per test taker','Percent in top 20% of HS class','Percent aspiring to more than BA','Percent family income more than $35,000','Percent both parents with BA'),
  local = c(550,770,2.8,27.8,30.6,11.2,9.0),
  in_state = c(1199,850,3.5,26.1,41.5,20.6,16.3),
  regional = c(1664,970,4.8,44.7,54.5,41.9,34.0),
  national = c(3766,980,5.3,45.7,62.2,43.0,37.1)
)

kable(
  x = market_segment,
  col.names = c('Characteristic','Local','In-state','Regional','National'),
  #caption = 'Simplified market segment profile\\\\ Connecticut Market 3: Fairfield County'
)
```

```{r sample-institutional-profile, echo = F}
#| label: tbl-sample-institutional-profile
#| tbl-cap: 'Simplified sample institutional profile for anonymous institution, students from Connecticut Market 3: Fairfield County'

sample_inst <- data.frame(
  characteristic = c('Total number of scores received','Pct of all test-takers in segment','lib_arts_num','lib_arts_share','engineering_num','engineering_share'),
  local = c(1,0.2,0,0.0,0,0.0),
  in_state = c(58,4.8,25,8.4,2,3.0),
  regional= c(69,4.1,35,7.0,5,5.8),
  national = c(109,2.9,61,5.3,3,0.8),
  total = c(237,3.3,121,5.9,10,1.8)
)

kable(
  x = sample_inst,
  col.names = c('','Local','In-state','Regional','National','Total'),
  #caption = 'Simplified sample institutional profile\\\\ Connecticut Market 3: Fairfield County'
)

>>>>

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```
The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. The [Online Appendix](https://structuring-college-access.github.io/) describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family income greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

[^5]: For example, the Market Segment Profile in the [Online Appendix](https://structuring-college-access.github.io/#/market-segment-profile) shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.

[^6]: The [Online Appendix](https://structuring-college-access.github.io/#/institutional-profile) reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer colleges, information that can be discerned from the Institutional Profiles (see [Online Appendix](https://structuring-college-access.github.io/#/institutional-profile)). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending behavior. Drawing from @RN4975, commodification of the Market Segment Model and Geomarkets engineers homophily.

@@

[^10]: Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket." In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago, Lake Forest College, and Saint Mary's College. <!--\[ADD TO APPENDIX CRYSTAL\]-->

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings [@RN4011]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to make generalizations about Geomarket variation nationally, but to explore how different geographical models of metropolitan areas, shaped by their unique demographic characteristics, influence the composition of prospect pools. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomena in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics. The [Online Appendix](https://structuring-college-access.github.io/#/results) presents results for a larger number of metropolitan areas.

@@

Results are presented below for Chicago, Dallas, and Los Angeles. The [Online Appendix](https://structuring-college-access.github.io/#/results) provides a more convenient means of viewing results. The online appendix also contains interactive maps, additional figures and tables not contained in the main text, and results for additional metropolitan areas (e.g., Bay Area, Philadelphia). The table of contents for navigating between metro areas can be accessed by clicking on the three horizontal lines on the bottom left of the screen.

@@

The interactive maps in the [Online Appendix](https://structuring-college-access.github.io/#/research-question-1-15) can show these shifts at the census tract-level. For example, in 1980 Black people were concentrated in southern tracts of the City of Chicago Geomarket, as described by scholarship on gentrification in Chicago [@ewing2024_nofuture; @snidal2022_exodus]. However, in 2000 and 2020 Black populations shift towards northeast tracts of the South and Southwest Suburbs Geomarket

@@

These contribution patterns by race/ethnicity demonstrate that student list orders targeting prospects in the Chicago metropolitan area using Geomarket filters that exclude the City of Chicago -- as observed in some orders collected via our public requests -- result in the loss of substantially disproportionate shares of Black and Hispanic students from the resulting prospect pools. The [Online Appendix](https://structuring-college-access.github.io/#/results) shows that this pattern holds across both lower-range (1020-1150) and higher-range (1310-1600) SAT score filters. In both cases, the City of Chicago contributes disproportionately larger shares of Black and Hispanic prospects compared to all other Geomarkets. <!--# [REVISE THESE SENTENCES TO HAMMER THE LEDE -- IF WE EXCLUDE CITY OF CHICAGO GEOMAREKT, THE WAY THE ASU ORDER DOES, WE LOSE A DISPROPORTIONATE SHARE OF BLACK AND HISPANIC STUDENTS] -->

@@

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles (11 Geomarkets) -- results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), results for which are shown in the [Online Appendix](https://structuring-college-access.github.io/#/results). The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core [@Harris1945]. As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up of the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernardino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

@@

The Market Segment Model begins with the status attainment model idea that student demand for college is a function of parental education and parental income. @RN4982 illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that households of specific class characteristics and college aspirations are likely to live in particular communities and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile ([Online Appendix](https://structuring-college-access.github.io/#/market-segment-profile)), a standard output in EPS software. When considering new Geomarkets to target, EPS recommends colleges utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile ([Online Appendix](https://structuring-college-access.github.io/#/institutional-profile)).

@@
